\textbf{Adaptation of the PHQ-4 Short Screening for Depression and Anxiety to increase its Sensitivity to Subclinical Variability}

### 2023-01-09 analysis.Rmd
scoring function

df <- read.csv("https://raw.githubusercontent.com/RealityBending/IllusionGameReliability/main/data/preprocessed_questionnaires.csv") |>
# df <- read.csv("../IllusionGameReliability/data/preprocessed_questionnaires.csv") |> 
  select(Sex, Age, starts_with("Item_PHQ4")) |> 
  datawizard::remove_empty_rows()

@@

>>>>

df <- read.csv("https://raw.githubusercontent.com/RealityBending/IllusionGameReliability/main/data/preprocessed_questionnaires.csv") |>
# df <- read.csv("../IllusionGameReliability/data/preprocessed_questionnaires.csv") |> 
  select(Sex, Age, starts_with("Item_PHQ4")) |> 
  filter(!is.na(Item_PHQ4_Anxiety_1))

@@

# Scores

```{r message=FALSE, warning=FALSE}
data <- sapply(cbind(df_anx, df_dep), function(x) {
  case_when(
      x == 0 ~ "Not at all",
      x == 1 ~ "Once or twice",
      x == 2 ~ "Several days",
      x == 3 ~ "More than half the days",
      TRUE ~ "Nearly every day"
    )
}) |> 
  as.data.frame()

source("score_PHQ4.R")

data <- score_PHQ4(A1=data$Anxiety_1, A2=data$Anxiety_2, D1=data$Depression_3, D2=data$Depression_4, method="basic") |> 
  datawizard::data_addsuffix("_Basic") |> 
  cbind(
    score_PHQ4(A1=data$Anxiety_1, A2=data$Anxiety_2, D1=data$Depression_3, D2=data$Depression_4, method="normalized") |> 
      datawizard::data_addsuffix("_Normalized")) |> 
  select(matches("Depression|Anxiety")) |> 
  pivot_longer(everything()) |> 
  separate(name, into = c("Dimension", "Scoring"))

data |> 
  ggplot(aes(x=value)) + 
  geom_area(data=estimate_density(data, at = c("Dimension", "Scoring"), method = "KernSmooth") |> 
              group_by(Scoring) |> 
              normalize(select = "y"),
            aes(x = x, y = y, fill=Dimension), alpha = 0.05, position="identity") +
  geom_hline(yintercept = 0.5, linetype = "dotted") +
  stat_ecdf(aes(color=Dimension), geom = "step", size=1) +
  facet_wrap(~Scoring, scales = "free") +
  scale_y_continuous(labels=scales::percent, expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("Depression" = "#3F51B5", "Anxiety" = "#E91E63"), guide = "none") +
  scale_color_manual(values = c("Depression" = "#3F51B5", "Anxiety" = "#E91E63")) +
  theme_modern(axis.title.space = 5) +
  labs(title = "Distribution and Cumulative Density of the Sample", x = "Score", y = "Cumulative Proportion")
```

### 2023-01-09 paper/manuscript.Rmd
scoring function

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
abstract: |
  The PHQ-4 is an ultra brief (4 items) screening questionnaire for depression and anxiety. We propose to add one aditional response option ("Once or twice", in between "Not at all" and "Several days") to improve its sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. Using Item Response Theory (IRT), we show that this new option does indeed capture specific portions of the measured constructs, and provide scoring procedure that preserve the compatability with past versions.

  **Significance Statement**. Developing reliable and sensitive instruments for mood disorders screening is critical in a global context marked by international crises (pandemics, wars), where more and more surveys as done online. In this study, we show that a small modification to the widely used PHQ-4 scale (adding the "Once or twice" response option) can increase its ability to capture the mild fluctuations prevalent in subclinical samples.

@@

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety [@kroenke2009ultra]. It consists of two items for depression [PHQ–2, @kroenke2003patient] and anxiety [GAD–2, @kroenke2007anxiety], each corresponding to DSM-V Diagnostic Criterion A symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations [@christodoulaki2022validation; @mendoza2022factor; @materu2020psychometric], becoming one of the most popular screening instruments for depression and anxiety [@maurer2018depression].

@@

## Original Scale

@@

## Revision

@@

The sample consists of 497 English-speaking participants (Mean age = 30.0, SD = 10.1, range: [18, 73]; Sex: 49.9\% females, 50.1\% males) from the general population who were administered the refined PHQ-4 online as part of another study, which data is available in open-access at https://github.com/RealityBending/IllusionGameReliability.

This study was approved by the NTU Institutional Review Board (NTU IRB-2022-187). All participants provided their informed consent prior to participation and were incentivized after completing the study.

@@

## Descriptive

@@

The objective of this study was to test the introduction of a "Once or twice" response option to the PHQ-4 to enhance its sensitivity to milder mood fluctuations. The fact that this new response option was the most prevalent response made by participants in our study is in itself evidence for its usefulness. IRT analysis further revealed that this response captures with precision a unique portion of anxiety and depression. Taken together, our results suggest that adding this option response increases the scale's potential to discriminate average mood levels (which are superior to zero) from lower-end extremes (the true zero). Future studies should further corroborate our statistical validation and test the usefulness of this revision on more targeted samples (e.g., clinical populations) and by cross-validating it with other more comprehensive measures.

@@

The dataset analysed during the current study are available in the GitHub repository https://github.com/DominiqueMakowski/PHQ4R

>>>>

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
abstract: |
  The PHQ-4 is an ultra brief (4 items) screening questionnaire for depression and anxiety. We propose to add one aditional response option ("Once or twice", in between "Not at all" and "Several days") to improve its sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. Using Item Response Theory (IRT), we show that this new option does indeed capture specific portions of the measured constructs, and provide a scoring procedure that preserve the compatability with past versions.

  **Significance Statement**. Developing reliable and sensitive instruments for mood disorders screening is critical in a global context marked by international crises (pandemics, wars), where increasingly more surveys are done online. In this study, we show that a small modification to the widely used PHQ-4 scale (adding the "Once or twice" response option) can increase its ability to capture the mild fluctuations prevalent in subclinical samples.

@@

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety [@kroenke2009ultra]. It consists of two items for depression [PHQ–2, @kroenke2003patient] and anxiety [GAD–2, @kroenke2007anxiety], each corresponding to DSM-5 diagnostic symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations [@christodoulaki2022validation; @mendoza2022factor; @materu2020psychometric], becoming one of the most popular screening instruments for depression and anxiety [@maurer2018depression].

@@

## Original Scale and Revision

@@

@@

The sample consists of 485 English-speaking participants (Mean age = 30.1, SD = 10.1, range: [18, 73]; Sex: 50.3\% females, 49.7\% males) from the general population who were administered the refined PHQ-4 online as part of another study, which data is available in open-access at https://github.com/RealityBending/IllusionGameReliability. This study was approved by the NTU Institutional Review Board (NTU IRB-2022-187). All participants provided their informed consent prior to participation and were incentivized after completing the study.

@@

## Descriptive Statistics

@@

The objective of this study was to test the introduction of a "Once or twice" response option to the PHQ-4 to enhance its sensitivity to milder mood fluctuations. The fact that this new response option was the most prevalent response made by participants in our study is in itself evidence for its usefulness. IRT analysis further revealed that this response captures with precision a unique portion of anxiety and depression. Taken together, our results suggest that adding this option response increases the scale's potential to discriminate average mood levels (which are superior to zero) from lower-end extremes (the true zero).

Future studies should further corroborate our statistical validation and test the usefulness of this revision on more targeted samples (e.g., clinical populations), cross-validate it with other more comprehensive measures, and refine the normalized scoring estimates in larger sample.

@@

The dataset analysed during the current study are available in the GitHub repository https://github.com/DominiqueMakowski/PHQ4R. We have created an R function for easy scoring according to the two methods described, which can be used as follows:

```{r message=FALSE, warning=FALSE}
# Load the function
source()
```
